The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-06 VegOzone.tex
Finished first draft of Sec. 3.1. No spell-check done.

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies peak in spring, with peak values of about $48\,\unit{ppb}$ in late April. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The ozone climatology laggs about 3 months behind the temperature climatology. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski can be explained due to the lower overall statistics.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is faily high (Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia at similar altitudes. In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for both, the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}) are shown. Darker colors indicate higher prohablity to observe these values. On top of the density distributions, 10 days averages of daily mean, daily maximum, and daily minimum \chem{[O_3]} are displayed together with the $1 \sigma$ uncertainties. For the daily mean the standard error of mean is shown in addition (albeit hardly visible). From the northern Fennoscandic climatology, one can deduce several features. The maximum of the 10 days average minimum ozone concentration peaks about $20\,\unit{days}$ earlier than the average mean and average maximum concentrations. From minimum concentration to maximum concentration it takes about $8.5\,\unit{months}$ while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of the average maximum ozone concentration is at $52\,\unit{ppb}$ (late April) and the minimum of the average minimum ocurres at $15\,\unit{ppb}$ (beginning of September). The steep rise of \chem{[O_3]} in spring coincident with increasing temperatures as well as increase in global irradiance.

@@

>>>>

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies peak in spring, with peak values of about $48\,\unit{ppb}$ in late April. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The ozone climatology laggs about 3 months behind the temperature climatology. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski can be explained due to the lower overall statistics.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia. In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher prohablity to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, daily maximum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}, and daily minimum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} are displayed together with the $1 \sigma$ uncertainties, respectively. In addition, \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} is also shown with standard error of mean (albeit small and hardly visible). Splines have been fitted through the data for easier intercomparison with ranalyis data.

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect of the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $8.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} ocurres at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring coincident with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ conseguatively (day of the year 110--150, Fig~\ref{fig:greening_season_change_Svanvik}).

As mentioned above, the average ozone concentration at Svanvik is in general lower than at the other sites which is in league with a decrease in tropospheric ozone with decreasing altitude. The densitiy distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. Only does the spring peak apparently occure slightly earlier (in the beginning of April).

@@

Finally, we shall look at the comparison of climatologies derived from observational data ozone and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modelled ozone concentrations in the north are slightly to low.
